---
title: "arplsbaseline_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{baseline_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(arplsbaseline)
```

# Background

It is often necessary to calculate a baseline for data, especially for spectral data, where a baseline can model the excessive noise in the spectrum. Noise-corrected data can be obtained by excluding that baseline of data from the original spectrum.

# Basic Baseline Function Call

To obtain a spectrum using the baseline function, the functions x and y must be vectors of some sort:

```{r}
x <- 1:200
y <- rchisq(200, df = 7)
```

# Using the Spectrum

The output of the baseline function is a Spectrum object, which has four extractable data elements and two methods associated with the class.

## Element Examples

```{r}
```

Importantly, the baseline and corrected data vectors can be viewed:

```{r}
#print("Baseline")
#head(spec$baseline)

#print("Corrected Data")
#head(spec$corrected)
```

## Method Examples

The print method can be applied to the Spectrum object:

```{r}
#print(spec)
```

A summary method with statistics about the corrected data can also be applied to the Spectrum object:

```{r}
#summary(spec)
```