---
title: "arplsbaseline_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{arplsbaseline_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "80%",
  fig.align = "center"
)
```

```{r setup}
library(arplsbaseline)
```

# Background

It is often necessary to calculate a baseline especially for spectroscopic data, where a baseline can model the excessive noise in the spectrum. Noise-corrected data can be obtained by subtracting that baseline of data from the original spectrum intensities. This baseline correction uses the Asymmetrically Reweighted Penalised Least Squares Smoothing (ARPLS) method @baek2015 and further information about how this method works can be found in the Theory Vignette.

# Basic Baseline Function Call

To obtain a spectrum using the baseline function, the baseline function must use a dataframe, with two columns of data.

The package contains some sample data with three different spectra (ham, redwine, strawberry). The data can be viewed calling one of these three names:

```{r}
# strawberry

# here only the top of the dataframe is shown to save space
head(strawberry)
```

and loaded into the environment by:

```{r}
data(strawberry)
```

Creating a spectrum with default parameter values can be done by the baseline function taking the dataframe as a sole argument.

```{r}
spec <- baseline(strawberry)
```

Running the function allows the user to also see how long it took the results to converge, in seconds, and in terms of many iterations of the algorithm was needed.

# Using the Spectrum

Now that a spectrum is obtained, we can extract its elements and call functions on it.

An overloaded print method can be applied to the Spectrum object which tells the user what type of object it is, how many signals were in the spectrum as well as what elements can be extracted.

```{r}
print(spec)
```

Another option is to just use the built-in method from R utils to view the structure of the object.

```{r}
str(spec)
```

## Element Examples

It can be shown that what the baseline function returns is in fact of `Spectrum` class, and that there are four extractable elements, all with same length that are associated with the object.

The original data points can be extracted:

```{r}
print("Wavenumbers")
head(spec$wavenumber)

print("Signal Intensity")
head(spec$original_signal)
```

Importantly, the baseline and corrected data vectors can be extracted:

```{r}
print("Baseline")
head(spec$baseline)

print("Corrected Data")
head(spec$corrected_signal)
```

## Method Examples

Asides from the print method, a summary method with basic statistics about the corrected data can also be applied to the object:

```{r}
summary(spec)
```

To visualise the spectra obtained, a plot method can be called which will show the original data, the calculated baseline and the newly corrected data.

```{r}
plot(spec)
```

A coercion method to a dataframe is also available so that instead of extracting vectors from the spectrum, all the data is available in one place.

```{r}
head(as.data.frame(spec))
```

A slightly more complicated method for the Spectrum object is a method that fits a model to the predicted baseline.

```{r}
# if not wanting the full summary, and the model returned
#baseline_gam(spec)

baseline_model <- baseline_gam(spec, full_summary = TRUE, return_gam = TRUE)
```

What is shown first, is a basic summary of the GAM (Generalised Additive Model) which is fit to the baseline. If the full_summary argument is set to TRUE, the full detailed summary output is printed as well to the console. There is an additional return_gam argument, which if set to TRUE, then the function will return the fitted GAM model.

GAMs are generalised linear models in which the linear response (signal intensity) depends on smooth functions of the predictors (wavenumbers). These are very flexible models, so estimating the form of the baseline equation using one of these models, rather than a high degree polynomial was preferred.

Visualisations of the fitted GAM with the original baseline, as well as the residuals from the GAM model are also presented by calling this function.

# Modifying Baseline Parameter

Obtaining a spectrum is relatively easy, but the spectrum may not look like how we want it to - there can be too much noise left in the spectrum, or a baseline has been fitted too closely. This is where the lambda parameter comes in.

We can change lambda to be bigger or smaller to our liking, as long as lambda is not less than 0. The default lambda is currently 10000.

## Decreasing Lambda

```{r}
spec2 <- baseline(strawberry, lambda = 1) 
```

```{r}
summary(spec2)
plot(spec2)
```

Decreasing lambda shows that the baseline follows the original data much closer, which squashes more of the corrected peaks.

The ratio parameter controls how much precision is wanted when comparing baselines to each other in order to determine when to stop the algorithm. Smaller ratios may result in less accurate predictions, but will be faster in cases where we are using datasets with lots of samples, and will mean that the algorithm definitely converges.

## Increasing Lambda

```{r}
spec3 <- baseline(strawberry, lambda = 1e8) 
plot(spec3)

```

Here increasing lambda shows that now the baseline is just a straight line instead, and has become much more generalised.

The choice of lambda is a difficult choice but ultimately is up to the user.

# Masking Functionality

Asides from changing lambda, the user also has the option to set a mask for a range of wavelengths, so that the ARPLS algorithm will ignore these, in computation.

If I wanted to mask the wavelengths from $1500 - 1800 \ cm^{-1}$:

```{r}
spec4 <- baseline(strawberry, start_mask = 1500, end_mask = 1800)
plot(spec4)
```

As a result, the baseline for these wavelengths have been calculated as 0, and the corrected signal has remained unchanged from the original. This has overall created a higher baseline for the rest of the spectrum, and values are much more corrected, than without the masking. Masking functionality may be good in situations where a specific section of the spectrum is known, where it is not very useful to use these values to compute a baseline.

# References
